---
title: "Learn package `rtweet`"
author: "Leo Williams"
date: "2021 Jan 17"
output: 
  bookdown::pdf_document2:
  	pandoc_args:
  	  - --pdf-engine=xelatex
---

```{r setup}
library(tidyverse)
library(rtweet)

# set ggplot theme
theme_set(
	theme_classic() +
	theme(
		text = element_text(family = "Courier", face = "bold")
	)
)
```

# Vignette `intro`

```{r}
# vignette("intro", package = "rtweet")
```

## Searching tweets

Search for 18,000 tweets containing the hashtag $\left (\frac{x^2sd}{\log(y)}\right )$ `#rstats`

```{r}
rt = search_tweets(q = "#rstats", 
					n = 1000, 
					include_rts = F, retryonratelimit = T)
head(rt)
```

See how many tweets from each account

```{r}
counts = rt %>%
	group_by(screen_name) %>%
	count() %>%
	arrange(desc(n))
head(counts)
```

Plot the counts by account name

```{r}
counts %>%
	filter(n > 30) %>%
	ggplot(aes(
		x = reorder(screen_name, n),
		y = n
	)) + 
	geom_col(color = "black", fill = "gray") +
	coord_flip() +
	labs(
		title = "\"#rstats\" tweets per account",
		x = "Twitter name", 
		y = "Tweets with \"#rstats\""
	)
```

Preview users' data:

```{r}
users_data(tweets = rt)
```

Plot time series for tweet search results

```{r}
ts_plot(data = rt) +
	labs(x = "Time", y = "Number of tweets")
```

Quickly visualize frequency of tweets over time using `ts_plot()`

```{r}
ts_plot(data = rt, by = "3 hours") +
	labs(
		title = "Frequency of #rstats Twitter statuses from past 9 days"
	)
```

# Stream tweets

# Get friends

Get a list of all the accounts a particular user follows:

```{r}
# get user IDs of accounts followed by CNN
cnn_fds = get_friends("cnn")

# look up data on those accounts
cnn_fds_data = lookup_users(cnn_fds$user_id)
```

```{r}
leo_fds = get_friends(users = "Leo Williams") ; head(leo_fds)
leo_fds_data = lookup_users(leo_fds$user_id) ; head(leo_fds_data)
```

# Get favorites

```{r}
jkr <- get_favorites("jk_rowling", n = 3000) ; head(jkr)
```

# Search users

```{r}
search_users("Trump", n = 1000) %>% head()
```

# Get trends

Discover what's currently trending in New Orleans

```{r}
get_trends("new orleans") %>% head()
```

# Look up users

lookup users by screen_name or user_id

```{r}
users <- c("KimKardashian", "justinbieber", "taylorswift13",
           "espn", "JoelEmbiid", "cstonehoops", "KUHoops",
           "upshotnyt", "fivethirtyeight", "hadleywickham",
           "cnn", "foxnews", "msnbc", "maddow", "seanhannity",
           "potus", "epa", "hillaryclinton", "realdonaldtrump",
           "natesilver538", "ezraklein", "annecoulter")
famous_tweeters <- lookup_users(users)
```

preview users data

```{r}
famous_tweeters %>% head()
```

extract most recent tweets data from the famous tweeters

```{r}
tweets_data(famous_tweeters) %>% head()
```

# Posting statuses

Post a status

```{r}
# post_tweet("Testing out sending a tweet using the #rtweet package #rstats")
```

# Following users

Follow users based on their screen name:

```{r}
# post_follow(user = "kearneymw")
```
