---
title: '`purrr` Tutorial'
author: 'Leo Williams'
date: |
    | Created: 2020 05_May 11
    | Updated: `r format(Sys.time(), "%Y %m_%b %d")`
output: 
    html_notebook:
        toc: yes
        toc_float: yes
params:
    IN_FILE: NULL
    # string -- path to file
    IN_DIR: '.'
    # string -- path to dir
    OUT_DIR: NULL
    # string -- path to dir
    OUT_OPT_ID: NULL 
    # string
    ## examples
    # pstring: "hello"
    # pvector: !r c("hi", "hello", "heya")
    # plist: !r list(A = c(a = "one", b = "two"), B = c(c = "three", d = "four", e = "five"))
    # p5bool: yes
    # p4null: NULL
---

# About
Adapted from the `purrr` tutorial by Charlotte Wickham at [Github](https://github.com/cwickham/purrr-tutorial)

FILENAME:
AUTHOR:
DATE:
USAGE:
DESCRIPTION:
VERSION:
NOTES:
R VERSION:
DEV PLATFORM:

# Setup

```{r setup}

# 1. load packages ----
library(knitr)
library(tidyverse)
library(repurrrsive)

# 2. set global knit options ----
opts_knit$set(
    # set project root dir as working dir
    root.dir = rprojroot::find_rstudio_root_file()
)

# 3. Set working dir for setup chunk ----
setwd(rprojroot::find_rstudio_root_file())
getwd()

# 4. Make results dir ----
# if (!is.null(params[["out_opt_id"]])) {
#     dir_results = paste0(params$OUT_DIR,
#                          "/results",
#                          "--",
#                          params$OUT_OPT_ID,
#                          as.character(format(Sys.time(),
#                                              '--%Y-%m-%d-%H%M%S')))
#     dir.create(dir_results)
# } else {
#     dir_results = paste0(params$OUT_DIR,
#                          "/results",
#                          as.character(format(Sys.time(),
#                                              '--%Y-%m-%d-%H%M%S')))
#     dir.create(dir_results)
# }
# 
# # * 4.1. make tmp and results for tables and plots subdirectories ----
# dir_results_tables = paste0(dir_results, "/tables")
# dir.create(dir_results_tables)
# dir_results_plots = paste0(dir_results, "/plots")
# dir.create(dir_results_plots)
# 
# dir_tmp = paste0(dir_results, "/tmp")
# dir.create(dir_tmp)
# 
# # 5. get filepath of current script to copy into results dir ----
# src_filepath = rstudioapi::getSourceEditorContext()$path
# file.copy(from = src_filepath, to = dir_results)
# 
# # 6. Export shell environment variables ----
# Sys.setenv(
# 	# results dirs and subdirs
# 	dir_results = dir_results, 
# 	dir_results_plots = dir_results_plots, 
# 	dir_results_tables = dir_results_tables, 
# 	# tmp dir
# 	dir_tmp = dir_tmp,
# 	# filepath of current script
# 	src_filepath = src_filepath
# )
# 
# # 7. Set global chunk options ----
# opts_chunk$set(
# 	tidy.opts = list(width.cutoff = 60),
# 	tidy = TRUE,
# 	dev = "pdf",
# 	fig.path = paste0(dir_results, "/figures-Rmd/")
# )
# 
# # 8. set ggplot theme ----
# theme_set(theme_classic() +
#           theme(panel.grid.major = element_line(colour = "#DDDDDD"),
#                 panel.grid.minor = element_line(color = "#FFFFFF"),
#                 text = element_text(size = 16, face = "bold")))
```

# Data
Data for Star Wars objects are loaded with `repurrrsive`.

## Inspect `sw_people`
How many elements are in `sw_people`?
```{r}
length(sw_people)
```

Who is the first person in `sw_people`?
```{r}
sw_people[[1]]$name
```

What's the difference between `sw_people[1]` and `sw_people[[1]]`?
```{r}
str(sw_people[1])
class(sw_people[1])
# list of a list

str(sw_people[[1]])
class(sw_people[[1]])
# just the list
```

# `map()`
`map(.x, .f, ...)` means: for each element of `.x`, do `.f`
- `.x` : a vector, list, or data frame
- `.f` : a function

How many starships has each character been in?
- For each person in `sw_people`, count the number of starships

**Strategy**:
- do it for one element
- turn it into a recipe
- use `map()` to do it for all elements

## 1. Do it for one element
Do it just for the first character, Luke Skywalker:
```{r}
luke <- sw_people[[1]]
length(luke$starships)
```

Another character, Princess Leia:
```{r}
leia <- sw_people[[5]]
length(leia$starships)
```

## 2. Turn it into a recipe
Reformat to `map()` input:
```{r}
luke <- sw_people[[1]]
length(luke$starships) # -->

# ____ <- sw_people[[____]]
# length(____$starships) # -->

# name <- list[[element]]
# fun(name$element2) # -->

# map(list, ~ fun(.x$element2)), where .x is iterates over every element

```
`~` is the formula, `.x` is the "pronoun" for one element of the input vector (e.g. one *list* per character from the input vector of lists)

## 3. Use `map()` to do it for all elements
```{r}
map(sw_people, ~ length(.x$starships)) %>% head()
```

## Another example
Find the name of each character's homeworld.

Create `planet_lookup` (ignore details for now):
```{r}
planet_lookup <- map_chr(sw_planets, "name") %>%
    set_names(map_chr(sw_planets, "url"))
planet_lookup
```


# Session Info

```{r save session data}
# # save data image
# save.image(file = paste0(dir_results, "/data.Rdata"))
# 
# # save session info
# write_lines(capture.output(sessionInfo()), 
#             path = paste0(dir_results, "/sessionInfo.txt"))
```
