---
title: "Learn: IRanges"
output:
  html_notebook:
    toc: true
    toc_float: true
---

Adapted from "An Overview of the *IRanges* package" (Aboyoun, et. al 2019) [vignette](http://bioconductor.org/packages/release/bioc/html/IRanges.html).

# Setup

```{r setup, results='hide'}
library(IRanges)
```

# IRanges objects

To make an IRanges object, call the `IRanges()` function. Ranges are normally specified by passing 2 of the 3 parameters: start, end, and width. All of these construct the same type of object.

```{r}
ir1 = IRanges(start = 1:10, width = 10:1)
ir1
str(ir1)
```

```{r}
ir2 = IRanges(start = 1:10, end = 11)
ir2
ir3 = IRanges(end = 11, width = 10:1)
ir3
identical(ir1, ir2) && identical(ir1, ir3)
```

```{r}
ir = IRanges(c(1, 8, 14, 15, 19, 34, 40), 
			 width = c(12, 6, 6, 15, 6, 2, 7))
ir
```

Access the starts, ends, and widths with the `start()`, `end()`, and `width()` functions.

```{r}
start(ir)
end(ir)
width(ir)
```

Subsetting an IRanges object is supported by numerical and logical indices.

```{r}
ir[1:4]
start(ir)
start(ir) <= 15
ir[start(ir) <= 15]
```

To illustrate range operations, here's a function to plot ranges:

```{r}
plotRanges = function(x, xlim=x, main=deparse(substitute(x)), col="black", sep=0.5, ...) {
	height = 1
	if (is(xlim, "IntegerRanges"))
		xlim = c(min(start(xlim)), max(end(xlim)))
		bins <- disjointBins(IRanges(start(x), end(x) + 1))
		plot.new()
		plot.window(xlim, c(0, max(bins) * (height + sep)))
		ybottom <- bins * (sep + height) - height
		rect(start(x)-0.5, ybottom, end(x)+0.5, ybottom + height, col=col, ...) 
		title(main) 
		axis(1)
}

plotRanges(ir)
```

[got too complicated...]