---
title: 'Learn R packge TCGAbiolinks'
author: 'Leo Williams'
output: html_notebook
---

# ABOUT

This notebook is for learning the R Bioconductor package `TCGAbiolinks`. It follows the vignettes from the package's [webpage](https://bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html).

# SETUP

Download required libraries (from [Introduction](#introduction))

```{r setup, message=FALSE}
library(TCGAbiolinks)
library(dplyr)
library(DT)
```

# 1. INTRODUCTION

[Vignette](https://bioconductor.org/packages/release/bioc/vignettes/TCGAbiolinks/inst/doc/index.html)

TCGAbiolinks is able to access The National Cancer Institute (NCI) Genomic Data Commons (GDC) thorough its
[GDC Application Programming Interface (API)](https://gdc.cancer.gov/developers/gdc-application-programming-interface-api) to search, download and prepare relevant data for analysis in R.

# 2. SEARCHING GDC DATABASE

## Useful information

Vignette: [TCGAbiolinks: Searching GDC database](https://bioconductor.org/packages/release/bioc/vignettes/TCGAbiolinks/inst/doc/query.html)

There are 2 available sources to download GDC data using `TCGAbiolinks`:

- GDC Legacy Archive
- GDC harmonized database

Information on understanding TCGA barcodes: [GDC TCGA barcodes](https://docs.gdc.cancer.gov/Encyclopedia/pages/TCGA_Barcode/)

## Searching arguments

Search *through* GDC data with the `GDCquery()` function. The function works well with the ***following*** arguments:

- `data.category`: valid project

```{r view projects}
# view available projects
gdc_projects = getGDCprojects()
gdc_projects %>% glimpse()

# view project information
TCGAbiolinks:::getProjectSummary("TCGA-READ")
```

