---
title: "Practice: Intuitive Biostatistics"
subtitle: "Nonlinear Regression"
---

# SETUP

```{r setup, message=F}
library(tidyverse)
library(extrafont)
library(drc)
library(nlme)
library(ggpmisc)

theme_set(
	theme_classic() +
	theme(
		text = element_text(family = "Anonymous Pro", face = "bold", size = 12)
	)
)
```

# CHAPTER 36: NONLINEAR REGRESSION

## GET DATA

Sample data: Bladder muscle relaxation data for one young rat

```{r nls-data}
dat_rat = tibble(
	log_norepinephrine_M = c(
		-8,-7.5,-7,-6.5,-6,-5.5,-5,-4.5,-4,-3.5,-3
	), 
	pct_relaxation = c(
		2.6,10.5,15.8,21.1,36.8,57.9,73.7,89.5,94.7,100,100
	)
)
glimpse(dat_rat)
```

## VISUALIZE

```{r}
dat_rat %>%
	ggplot(aes(
		x = log_norepinephrine_M, 
		y = pct_relaxation
	)) + 
	geom_point() +
	labs(
		title = "Effect of norepinephrine on bladder muscle relaxation", 
		x = "Norepinephrine concentration (log)",
		y = "% Muscle relaxation"
	)
```

# MODEL

## Test nls

Test `nls()` from <https://www.r-bloggers.com/2016/02/first-steps-with-non-linear-regression-in-r/> using the following equation:

$$
y = \frac{a \cdot x}{b + x}
$$

```{r test-nls}
#simulate some data
set.seed(20160227)
x <- seq(0,50,1)
y <- ((runif(1,10,20)*x)/(runif(1,0,10) + x)) + rnorm(51,0,1)
test_dat = tibble(x, y)
glimpse(test_dat)
#for simple models nls find good starting values for the parameters even if it throw a warning
m <- nls(y ~ a*x/(b + x), data = test_dat)
# summarize model
summary(m)
#get some estimation of goodness of fit
cor(y,predict(m))
# [1] 0.9496598
#plot
plot(x,y)
lines(x,predict(m), lty = 2, col = "red", lwd = 3)
```


## Dose-response curve

The model used is a standard model for a drug dose-response relationship:

$$
Y = Bottom + \frac{Top - Bottom}{1 + 10^{(LogEC_50 - X) \cdot HillSlope}} \\
Y = A + \frac{B - A}{1 + 10^{(C - X) \cdot D}}
$$

Model this nonlinear relationship using the sample data, where: - a = Bottom - b = Top - c = log(EC_50) - d = HillSlope

```{r model}
# mod_dr = nls(formula = y ~ b + (t - b)/(1 + 10^((e - x) * s)), 
# b=0
mod_dr = nls(formula = pct_relaxation ~ 0 + (t - 0)/(1 + 10^((e - log_norepinephrine_M) * s)),
#  = nls(formula = pct_relaxation ~ t/(1 + 10^((e - log_norepinephrine_M) * s)),
			   data = dat_rat, 
			   start = list(
			   	t = 100, 
			   	e = -5.5,
			   	s = 0.5
			   ))
summary(mod_dr)
dat_rat$preds = predict(mod_dr)
```

plot the model

```{r}
dat_rat %>%
	ggplot(aes(
		x = log_norepinephrine_M, 
		y = pct_relaxation
	)) + 
	geom_point() +
	geom_point(aes(y = preds), color = "red") +
	geom_smooth(method = 'nls', 
				method.args = list(
					formula = y ~ 0 + (t - 0)/(1 + 10^((e - x) * s)),
			   # data = dat_rat, 
			   start = list(
			   	t = 100, 
			   	e = -5.5,
			   	s = 0.5
			   )), 
				se = F, size = 0.5, linetype = "dashed") +
	labs(
		title = "Effect of norepinephrine on bladder muscle relaxation", 
		x = "Norepinephrine concentration (log)",
		y = "% Muscle relaxation"
	) 
```

## Michaelis-Menten

Michaelis-Menten equation:

$$
Y = \frac{V_{max} \cdot X}{K_m + X}
$$

Which equates to the function:

$$
f(x, \beta) = \frac{\beta_1 x}{\beta_2 + x}
$$

It's nonlinear because it cannot be expressed as a linear combination of the two $\beta$.

Load some sample MM data from the blog post [Plotting and fitting enzymology data... \| R for Biochemists](https://rforbiochemists.blogspot.com/2015/05/plotting-and-fitting-enzymology-data.html)

```{r sample-mm-data}
# As a biochemist, one of the first data types we play with is enzymology data. 
# Here is an example of some enzyme simple kinetic data
# Eight values for substrate concentration with corresponding enzyme velocities. 
# The goals are to plot the data, in different formats, and also, find the best fit values of Km and Vmax
# all of this is complete in base R, so no packages should be needed

# Entered as vectors 
mm_data = tibble(
	S = c(0,1,2,5,8,12,30,50),
	V = c(0,11.1,25.4,44.8,54.5,58.2,72.0,60.1)
)
glimpse(mm_data)
```

Plot the sample data:

```{r plot-mm-data}
mm_data %>%
	ggplot(aes(
		x = S, 
		y = V
	)) +
	geom_point() +
	geom_smooth(size = 0.5, alpha = 0.1, method = "lm", se = F) +
	labs(
		title = "Michaelis-Menten Kinetics",
		x = "Substrate Concentration", 
		y = "Enzyme Velocity"
	)
```

Try and model a michaelis menten curve to fit the enzyme data:

```{r nls-mm}
mm_form <- V ~ (Vmax*S)/(Km + S)
mod_mm = nls(formula = mm_form, data = mm_data, start = list(Vmax = 70, Km = 5))
summary(mod_mm)
confint(mod_mm)
mod_mm_pred = as_tibble(predict(object = mod_mm, interval = "confidence", newdata = mm_data$S, level = 0.95))
mod_mm_pred
profile_mod_mm = profile(mod_mm)
profile_mod_mm
plot(profile_mod_mm)
```

Predict new values with `predict()`:

```{r mm-predict}
mm_data$predict = predict(mod_mm)
glimpse(mm_data)
```

Plot the model fit onto the raw data:

```{r}
# plot(mm_data$S, mm_data$V)
# plot(mod_mm, log = '', pch = 17, main = "Fitted MM curve")
# plot(mod_mm)
```

```{r plot-mm-fit}
mm_data %>%
	ggplot(aes(
		x = S, 
		y = V
	)) +
	geom_point() +
	geom_point(aes(y = predict), color = "red") +
	# add incorrect linear regression line for reference
	# geom_smooth(method = "lm", size = 0.5, se = F, color = "black") +
	geom_smooth(method = 'nls', 
				method.args = list(
					formula = y ~ (a*x)/(b + x), 
					start = list(a = 70, b = 5)
				), 
				se = F, size = 0.5, linetype = "dashed") +
	labs(
		title = "Michaelis-Menten NLS Fit", 
		y = "Enzyme Velocity (V)", 
		x = "Substrate Concentration (S)"
	)
	# geom_ribbon(data = mod_mm_pred, aes(x = S, ))
```
